---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ripserr: Calculate Persistent Homology of Vietoris-Rips and Cubical Complexes using Ripser in R

[![Travis-CI Build Status](https://travis-ci.org/rrrlw/ripserr.svg?branch=master)](https://travis-ci.org/rrrlw/ripserr)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/rrrlw/ripserr?branch=master&svg=true)](https://ci.appveyor.com/project/rrrlw/ripserr)
[![Codecov test coverage](https://codecov.io/gh/rrrlw/ripserr/branch/master/graph/badge.svg)](https://codecov.io/gh/rrrlw/ripserr?branch=master)

[![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
[![CRAN version](http://www.r-pkg.org/badges/version/ripserr)](https://CRAN.R-project.org/package=ripserr)
[![CRAN Downloads](http://cranlogs.r-pkg.org/badges/grand-total/ripserr)](https://CRAN.R-project.org/package=ripserr)

## Overview

ripserr ports the [Ripser](https://arxiv.org/1908.02518) and [Cubical Ripser](https://arxiv.org/abs/2005.12692) persistent homology calculation engines from C++ via [Rcpp](https://CRAN.R-project.org/package=Rcpp).
It can be used as a convenient and rapid calculation tool in topological data analysis pipelines.

## Installation

```{r install, eval=FALSE}
# install development version
devtools::install_github("rrrlw/ripserr")

# install from CRAN (not available yet)
# install.packages("ripserr")
```

## Sample code

Ripser (Vietoris-Rips complex) can be used as follows for data with dimension greater than or equal to 2.

```{r sample-vr}
# load ripserr
library("ripserr")

set.seed(42)
SIZE <- 100

# 2-dimensional example
dataset2 <- rnorm(SIZE * 2)
dim(dataset2) <- c(SIZE, 2)
vr_phom2 <- vietoris_rips(dataset2)
head(vr_phom2)
tail(vr_phom2)

# 3-dimensional example
dataset3 <- rnorm(SIZE * 3)
dim(dataset3) <- c(SIZE, 3)
vr_phom3 <- vietoris_rips(dataset3, dim = 2) # default: dim = 1
head(vr_phom3)
tail(vr_phom3)
```

Cubical Ripser (cubical complex) can be used as follows for data with dimension equal to 2, 3, or 4.

```{r sample-cub}
# load ripserr
library("ripserr")

set.seed(42)
SIZE <- 10

# 2-dimensional example
dataset2 <- rnorm(SIZE ^ 2)
dim(dataset2) <- rep(SIZE, 2)
cub_phom2 <- cubical(dataset2)
head(cub_phom2)
tail(cub_phom2)

# 3-dimensional example
dataset3 <- rnorm(SIZE ^ 3)
dim(dataset3) <- rep(SIZE, 3)
cub_phom3 <- cubical(dataset3)
head(cub_phom3)
tail(cub_phom3)

# 4-dimensional example
dataset4 <- rnorm(SIZE ^ 4)
dim(dataset4) <- rep(SIZE, 4)
cub_phom4 <- cubical(dataset4)
head(cub_phom4)
tail(cub_phom4)
```

## Functionality

1. Calculation of persistent homology of Vietoris-Rips complexes using Ripser (function named `vietoris_rips`).
1. Calculation of persistent homology of cubical complexes using Cubical Ripser (function named `cubical`).

## Citation

If you use the ripserr package in your work, please consider citing the following (based on use):

* **General use of ripserr:** Wadhwa RR, Piekenbrock M, Scott JG. ripserr: Calculate Persistent Homology with Ripser-based Engines; version 0.1.0. URL https://github.com/rrrlw/ripserr.
* **Calculation using Vietoris-Rips complex:** Bauer U. Ripser: Efficient computation of Vietoris-Rips persistence barcodes. 2019; arXiv: 1908.02518.
* **Calculation using cubical complex:** Kaji S, Sudo T, Ahara K. Cubical Ripser: Software for computing persistent homology of image and volume data. 2020; arXiv: 2005.12692.

## Contribute

To contribute to ripserr, you can create issues for any bugs/suggestions on the [issues page](https://github.com/rrrlw/ripserr/issues).
You can also fork the ripserr repository and create pull requests to add useful features.