---
title: "Calculating Persistent Homology with a Cubical Complex"
author: "Raoul R. Wadhwa, Jacob G. Scott"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating Persistent Homology with a Cubical Complex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ripserr)
```

## Sample dataset

For this vignette, we will generate a lattice representing a 2-dimensional image; this will be stored in a variable named `sample_image`.

```{r load-data,fig.width=6,fig.height=4}
# create dataset
sample_image <- matrix(0, nrow = 10, ncol = 10)
i <- 2:9
j <- c(2, 9)
sample_image[i, j] <- 1
sample_image[j, i] <- 1

# view as matrix
sample_image

# view as image
graphics::image(sample_image, useRaster = TRUE, axes = FALSE)
```

Above, each of the 100 matrix values is analogous to a single pixel in an image.

## Calculating persistent homology

Based on the image, we expect a 1-cycle to be present in the persistent homology of `sample_image`.
The [CubicalRipser](https://arxiv.org/abs/2005.12692) C++ library is wrapped by R using [Rcpp](https://CRAN.R-project.org/package=Rcpp), and performs calculations via a cubical complex created with `sample_image`.
These calculations result in a data frame that characterizes the persistent homology of `sample_image` and can be performed with a single line of R code using ripserr.

```{r}
# calculate persistent homology
image_phom <- cubical(sample_image)

# print features
image_phom
```

Each row in `image_phom` represents a single feature.
The homology matrix has 3 columns:

1. **dimension:** if 0, represents a 0-cycle; if 1, represents a 1-cycle; and so on.
1. **birth:** radius of the cubical complex at which this feature begins
1. **death:** radius of the cubical complex at which this feature ends

Persistence of a feature is generally defined as the length of the interval of the radius within which the feature exists.
This is calculated as the numerical difference between the second (birth) and third (death) columns of the homology data frame
Confirmed in the output above, the homology data frame is ordered by dimension, with the birth column used to sort features of the same dimension.
As expected for `sample_image`, the homology data frame contains a single 1-cycle.
The [ggtda](https://CRAN.R-project.org/package=ggtda) and [TDAstats](https://CRAN.R-project.org/package=TDAstats) R packages can be used to visualize `image_phom` for further insight.
